The Chemistry Development Kit (CDK): atom typing, rendering, molecular formulas, and substructure searching

### 2016-09-29 article.tex
reworked figures to use \figure env

  The perception from coordinates and wedge assignment algorithms are fundamental for conversion
  between formats that store stereochemistry implicitly based on coordinates (e.g. molfile, CML) and 
  explicitly (e.g. SMILES, CML, InChI). Perception from 2D coordinates can optionally identify
  perspective projects, specifically: Fischer, Haworth, and Chair projections. With the perception of 
  perspective projections enabled, database entries currently consider distinct can be merged
  (Figure~3\ref{fig:stereoprojections}). The perception is based on an algorithm briefly 
  described by~\cite{Karapetyan2015}.

@@

  Many of the rendering options are available as parameters in the core API
  and as methods on the DepictionGenerator class. This includes substructure
  coloring, exemplified with an example reaction shown in Figure~4\ref{fig:depiction}.

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. Particularly, collision detection has been
  greatly improved. Figure~5\ref{fig:sdg} shows a difference in output between
  the old code base, with and without overlap resolving, and with the new code.
  The new code can be used like:

@@

CTfile Sgroups capture and organise high level information about sets of atoms
and bonds~\cite{Gushurst91}. There are four types of Sgroup: Display Shortcuts, Polymers,
Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
repeat units (e.g. bracketing) and abbreviations (Figure~8\ref{fig:sgroups}).
CDK~\cdkversion{} adds 
supports for representation, reading, writing, and depiction of Sgroups.

@@

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~4\ref{tab:modules} and the dependencies
between the CDK modules are depicted in Figure~6\ref{fig:deps}.

@@

\subsection*{Figure 2 - Relative storage of stereochemistry, the \textbf{type} and \textbf{focus} of stereochemsitry are fixed for a given stereocenter description but the \textbf{carriers} and \textbf{configuration} are relative.}\label{fig:stereodatastructure}
The multiple rows for each stereochemistry type are different internal representation that would be considered equivalent. In the tetrahedral types, hydrogens may be suppressed in a molecular graph so the \textbf{focus} is reused in the \textbf{carriers} list as a placeholder.

\includegraphics{img/stereodesc_annotated.pdf}

\subsection*{Figure 3 - The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).}\label{fig:stereoprojections}
Without perceiving the stereochemistry indicated by Haworth projection in CHEMBL23970, the database entries are incorrectly considered distinct. Down stream aggregation databases mirror this separation (PubChem CID 5280, CID 65119).

\includegraphics{img/projections_annotated.pdf}

\subsection*{Figure 4 - Integrated example showing the rendering and SMILES
parsing functionality. Example from U.S. Patent \textbf{US 2014 231770 A1} para 287.}\label{fig:depiction}

%% SMILES=[CH3:17][CH:16]([CH3:18])[CH2:15][Mg+].[cH:7]1[cH:6][cH:5][c:4]2[c:9]([cH:8]1)[cH:10][cH:11][c:2]3[c:3]2[n:12][cH:13][n:14][c:1]3Cl.[Br-]>C/C(=C/C(=O)C)/[O-].C/C(=C/C(=O)C)/[O-].C/C(=C/C(=O)C)/[O-].[Fe+3].CN1CCCC1=O.C1CCOC1.Cl>[CH3:17][CH:16]([CH3:18])[CH2:15][c:1]1[c:2]2[cH:11][cH:10][c:9]3[cH:8][cH:7][cH:6][cH:5][c:4]3[c:3]2[n:12][cH:13][n:14]1 |f:0.2,3.4.5.6,SgD::cdk&#58;ReactionConditions:16 h&#44; RT|
%% Kumada Coupling?

\includegraphics[width=\textwidth]{img/US20140231770A1_0287.pdf}

\subsection*{Figure 5 - The improved structure diagram generation has improved
code to solve overlap.}\label{fig:sdg}
The original SDG code used general heuristics (left) and the
OverlapResolver would fine tune the layout to ensure atoms would not be placed
at the same location (middle). The new SDG algorithm is able to
make more rigorous changes, making the final output must more pleasing
(right).

\includegraphics[width=\textwidth]{sdg.png}

\subsection*{Figure 6 - Structure diagram generation for structures with
double bond and tetrahedral stereochemistry.}\label{fig:sdgstereo}

\includegraphics[width=0.3\textwidth]{sdg2.png}

  \subsection*{Figure 7 - Dependencies between CDK modules.}\label{fig:deps}
      Visualization of the dependencies between CDK modules. For example,
      the cdk-core depends on the cdk-interfaces module. A few higher level
      modules have been left out: cdk-builder3dtools, cdk-legacy, and
      cdk-depict.

\includegraphics[width=\textwidth]{cdkDeps.png}

\subsection*{Figure 8 - Examples of Sgroups now captured by the CDK and encoded in molfiles and CXSMILES. a) Ethyl esterification fully expanded reaction. b) Using Sgroup abbreviations allows display short cuts and more compact depiction. c) An example of a structure repeat unit in \textit{DNA 5'-phosphate} (\textbf{CHEBI:4294})}
\label{fig:sgroups}

\includegraphics[width=\textwidth]{img/sgroups.pdf}

@@

      \subsection*{Table 4 - A selection of key CDK modules with major changes.}\label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Dependencies between modules are depicted in Figure~4\ref{fig:deps}.
  \vskip 1\baselineskip

>>>>

  The perception from coordinates and wedge assignment algorithms are fundamental for conversion
  between formats that store stereochemistry implicitly based on coordinates (e.g. molfile, CML) and 
  explicitly (e.g. SMILES, CML, InChI). Perception from 2D coordinates can optionally identify
  perspective projects, specifically: Fischer, Haworth, and Chair projections. With the perception of 
  perspective projections enabled, database entries currently consider distinct can be merged
  (Figure~\ref{fig:stereoprojections}).  The perception is based on an  algorithm briefly  
  described by~\cite{Karapetyan2015}.

@@

  Many of the rendering options are available as parameters in the core API
  and as methods on the DepictionGenerator class. This includes substructure
  coloring, exemplified with an example reaction shown in Figure~\ref{fig:depiction}.

@@

  The structure diagram layout has been rewritten and the new code solves a
  number of long standing issues. Particularly, collision detection has been
  greatly improved. Figure~\ref{fig:sdg} shows a difference in output
  between  
  the old code base, with and without overlap resolving, and with the new code.
  The new code can be used like:

@@

CTfile Sgroups capture and organise high level information about sets of atoms
and bonds~\cite{Gushurst91}. There are four types of Sgroup: Display Shortcuts, Polymers,
Mixtures, and Data. The most familiar Sgroups from an end user perspective are structure 
repeat units (e.g. bracketing) and abbreviations (Figure~\ref{fig:sgroups}).
CDK~\cdkversion{} adds 
supports for representation, reading, writing, and depiction of Sgroups.

@@

An overview of key modules with description, important changes, and dependencies
on third-party libraries is given in Table~4\ref{tab:modules} and the dependencies
between the CDK modules are depicted in Figure~\ref{fig:deps}.

@@

\begin{figure}[h!]
  \caption{\csentence{Relative storage of stereochemistry, the \textbf{type} and
    \textbf{focus} of stereochemsitry are fixed for a given
    stereocenter description but the \textbf{carriers} and
    \textbf{configuration} are relative.} 
  The multiple rows for each
  stereochemistry type are different internal representation that
  would be considered equivalent. In the tetrahedral types,
  hydrogens may be suppressed in a molecular graph so the
  \textbf{focus} is reused in the \textbf{carriers} list as a
  placeholder.}
\centering
\includegraphics{img/stereodesc_annotated.pdf}
\label{fig:stereodatastructure}
\end{figure}

\begin{figure}[h!]
  \caption{\csentence{The raw input files of CHEMBL23970 and CHEMBL444314 are displayed (ChEMBL 21).}
    Without perceiving the stereochemistry indicated by Haworth
    projection in CHEMBL23970, the database entries are incorrectly
    considered distinct. Down stream aggregation databases mirror this
    separation (PubChem CID 5280, CID 65119).}
    \centering
    \includegraphics{img/projections_annotated.pdf}
    \label{fig:stereoprojections}
\end{figure}

\begin{figure}[h!]
  \caption{\csentence{Integrated example showing the rendering and SMILES
      parsing functionality. Example from U.S. Patent \textbf{US 2014 231770 A1} para 287.}}
  \centering
  \includegraphics[width=\textwidth]{img/US20140231770A1_0287.pdf}
  \label{fig:depiction}
\end{figure}

\begin{figure}[h!]
  \caption{\csentence{{The improved structure diagram generation has improved
        code to solve overlap.}}
    The original SDG code used general heuristics (left) and the
    OverlapResolver would fine tune the layout to ensure atoms would not be placed
    at the same location (middle). The new SDG algorithm is able to
    make more rigorous changes, making the final output must more pleasing
    (right).}
    \includegraphics[width=\textwidth]{sdg.png}
    \label{fig:sdg}
\end{figure}

\begin{figure}[h!]
  \caption{\csentence{Structure diagram generation for structures with
      double bond and tetrahedral stereochemistry.}}
  \centering
  \includegraphics[width=0.3\textwidth]{sdg2.png}
  \label{fig:sdgstereo}
\end{figure}

\begin{figure}[h!]
  \caption{\csentence{Dependencies between CDK modules.}
    Visualization of the dependencies between CDK modules. For example,
    the cdk-core depends on the cdk-interfaces module. A few higher level
    modules have been left out: cdk-builder3dtools, cdk-legacy, and
    cdk-depict.}
  \centering
  \includegraphics[width=\textwidth]{cdkDeps.png}
  \label{fig:deps}
\end{figure}

\begin{figure}[h!]
  \caption{\csentence{Examples of Sgroups now captured by the CDK and
      encoded in molfiles and CXSMILES. a) Ethyl esterification fully
      expanded reaction. b) Using Sgroup abbreviations allows display
      short cuts and more compact depiction. c) An example of a
      structure repeat unit in \textit{DNA 5'-phosphate}
      (\textbf{CHEBI:4294})}}
  \includegraphics[width=\textwidth]{img/sgroups.pdf}
  \label{fig:sgroups}
\end{figure}

@@

      \subsection*{Table 4 - A selection of key CDK modules with major changes.}\label{tab:modules}
  An overview of a selection of often used CDK modules with description,
  dependencies on third-party libraries, and the major changes since
  version~1.2. Dependencies between modules are depicted in Figure~\ref{fig:deps}.
  \vskip 1\baselineskip
